---
output: html_document
title: "Log2FoldChange Analysis in mouse CD4-positive T cells"
author: "Mira Sohn" 
---


## Alignment & DE analysis

### - Alignment: [Link](https://github.com/Mira0507/mouse_cd4)

### - DE analysis: [Link I](https://github.com/Mira0507/mouse_cd4/blob/master/DE_EnsDb.Rmd), [Link II](https://github.com/Mira0507/mouse_cd4_reanalysis/blob/master/DE_EnsDb.Rmd)

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
    warning=FALSE,
    message=FALSE
    )
```

## Loading packages


```{r loading_packages}
library(data.table)
library(tidyverse)
library(rmarkdown)
library(ggplot2)
library(gridExtra)
library(ggrepel)
```

## Importing and cleaning ranking data

```{r data_import}

# Assign an input file name
filename <- "m.log2FoldChange_difference_SA_ST.csv"

# Import data
diff.high <- fread(filename)[1:50, -1][, Group := "High % Difference"]
diff.low <- tail(fread(filename), 50)[, -1][, Group := "Low % Difference"]

# Combine two datasets
diff <- rbind(diff.high, diff.low) %>%
    dplyr::select(-starts_with("Dfam"))

# Nest by Group
diff.nested <- diff %>% 
    group_by(Group) %>% 
    nest()



```

### Relationship between Difference & Percent

#### - Difference: Salmon - STAR 

#### - Percent: 100 x (Salmon - STAR) / (Salmon + STAR) 





```{r diff_and_percentdiff}


# Create a vector storing the first and last values 
vec.fn <- function(original.vec) { 

    len.of.vec <- length(original.vec)

    updated.vec <- c(round(original.vec[1], 5), 
                     rep("", len.of.vec-2),
                     round(original.vec[len.of.vec], 5)) 

    return(updated.vec)
}





# Set a function plotting relationship between difference and percent difference
plot1.fn <- function(df, title) {

    ggplot(df, aes(x=abs(Difference), 
                   y=abs(Percent))) + 

geom_point() + 
theme_bw() +
geom_smooth(method="lm", se=F) + 
scale_x_log10() + 
xlab("Absolute Difference (Salmon - STAR)") + 
ylab("100 x Absolute Difference / (Salmon + STAR)") + 
ggtitle(paste("Relationship between Absolute Difference and Absolute Percent Difference in", title, "Percent Difference Genes\n(with R-squared)")) +
geom_text(mapping=aes(x=Inf, y=Inf, label=diff.perdiff.rsq), 
          vjust=2, hjust=1.1, color="red", size=4) 

}


# Model linear regression, extract p-values, and unnest
diff.nested <- diff.nested %>%
    mutate(diff.perdiff.mod=map(data, ~ lm(abs(Percent) ~ abs(Difference), data=.x)),
           diff.perdiff.summ=map(diff.perdiff.mod, ~ summary(.x)),
           diff.perdiff.coef=map(diff.perdiff.summ, ~ .x$coefficients[2, 4]),
           diff.perdiff.rsq=unlist(map(diff.perdiff.summ, ~ .x$r.squared))) %>% 
    unnest(data, diff.perdiff.coef) 

# Explore the data
dim(diff.nested)
head(diff.nested)


# Add a column storing p-value for plotting
diff.nested$diff.perdiff.pval <- vec.fn(diff.nested$diff.perdiff.coef)

diff.nested$diff.perdiff.rsq <- vec.fn(diff.nested$diff.perdiff.rsq)

# Renest by Group and add plots
diff.nested <- diff.nested %>%
    group_by(Group) %>%
    nest() %>% 
    mutate(diff.perdiff.plot=map(data, ~ plot1.fn(.x, Group)))


# Print the plots
diff.nested$diff.perdiff.plot[[1]]
diff.nested$diff.perdiff.plot[[2]]

```


## Exploring Variables 



#### - Variant.Allels: Genetic Variation (e.g. SNPs) 

#### - Repetitive.Elements: found by [Tandem Repeats Finder](https://tandem.bu.edu/trf/trf.html) 

#### - **Tandem Repeats Finder**: "Tandem repeats finder: a program to analyze DNA sequences", Nucleic Acids Research, Volume 27, Issue 2, 1 January 1999, Pages 573–580



```{r explore_variables, fig.width=12, fig.height=10}


# Set a function calculating pvalue via two-tailed t-test
ttest.fn <- function(response.variable) {

    pval <- t.test(abs(response.variable) ~ Group, data=diff)$p.value
    pvec <- c(as.character(pval), rep("", 99))

    return(pvec)
}


# Create a vector storing the pvalues (character vector)
pval.vec <- c(ttest.fn(diff$Percent),
              ttest.fn(diff$Difference),
              ttest.fn(diff$num.trans), 
              ttest.fn(diff$Variants),
              ttest.fn(diff$TRF))


# Rename columns for plotting
metrics.rename <- c("Absolute.Difference", 
                    "Alternative.Transcripts", 
                    "Percent.Difference", 
                    "Variant.Alleles", 
                    "Repetitive.Elements")

colnames(diff) <- c("Metric", 
                    "GENEID", 
                    metrics.rename,
                    "Group") 



# Clean the data frame 
diff <- diff %>% 
    gather(Metrics, Value, -Metric, -GENEID, -Group)

metric.levels <- c("Percent.Difference",
                   "Absolute.Difference",
                   "Alternative.Transcripts",
                   "Variant.Alleles",
                   "Repetitive.Elements")

diff$Metrics <- factor(diff$Metrics, levels=metric.levels)


# Add a column storing pvalues
diff$metrics.pval <- pval.vec


# Explore the output after data cleaning
head(diff)
dim(diff)

# Plot 
ggplot(diff, aes(x=Group, y=abs(Value), color=Group)) + 
    geom_jitter(alpha=0.7) + 
    geom_boxplot(color="black", fill="grey", alpha=0) + 
    facet_wrap(~Metrics, scales="free_y") + 
    theme_bw() + 
    theme(strip.text.x=element_text(size=11),
          axis.text.x=element_text(size=10)) + 
    scale_y_log10() + 
    geom_text(aes(x=Inf, y=Inf, label=metrics.pval), 
              vjust=2, hjust=1.1, color="black", size=4) + 
ggtitle("Comparison between High vs Low Difference Genes in Diverse Variables\n(with P-Values from Two-Tailed T-Test)") + 
ylab("Value")






```



## Session info

```{r session_info}

sessionInfo()

```
