---
title: "TPM vs Count Inputs in Mouse CD4-Positive T cell RNA-seq" 
output: html_document
author: "Mira Sohn"
---

## Raw data 

#### - **GEO number**: [GSE128615](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE128615)

#### - **Study summary**: To determine the influence of differential Kdm6a expression in immune cells, whole transcriptome analysis for CD4+ T cells from WT and Kdm6a cKO mice were performed using RNA-Seq.

#### - **Reference**: [J Clin Invest. 2019 Aug 12;129(9):3852-3863. doi: 10.1172/JCI126250.](https://pubmed.ncbi.nlm.nih.gov/31403472)

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
    warning=FALSE,
    message=FALSE
    )
```



## Loading packages

```{r loading_packages}
library(data.table)
library(rmarkdown)
library(AnnotationHub)
library(tidyverse)
library(tximport)
library(ggplot2)
library(DESeq2)
library(gridExtra)
library(pheatmap)
library(UpSetR)
library(ensembldb)
library(apeglm)
library(ashr)
```

## Setting AnnotationHub


```{r annotationhub_setup}
DB <- "EnsDb"  # Assing your species
AnnotationSpecies <- "mus musculus"  # Assign your species 
ah <- AnnotationHub(hub=getAnnotationHubOption("URL"))   # Connect to annotation DB
```

## Running AnnotationHub 

```{r run_annotationhub}

# Filter annotation of interest
ahQuery <- query(ah, 
                 pattern=c(DB, AnnotationSpecies), 
                 ignore.case=T)      


# Select the most recent data
DBName <- mcols(ahQuery) %>%
    rownames() %>%
    tail(1)

AnnoDb <- ah[[DBName]] 

# Explore your EnsDb object with following accessors:
# columns(AnnpDb)
# keytypes(AnnoDb)
# keys(AnnoDb, keytype=..)
# select(AnnoDb, keys=.., columns=.., keytype=...)
AnnoKey <- keys(AnnoDb, keytype="TXID")

# Note: Annotation has to be done with not genome but transcripts 
AnnoDb <- select(AnnoDb, 
                 AnnoKey,
                 keytype="TXID",
                 columns=c("GENEID", "GENENAME")) 



# Check if your AnnoDb has been extracted and saved correctely
class(AnnoDb)
head(AnnoDb)
```


## Defining file name and path for .sf files

.sf files have been created from fastq data by salmon


```{r preparing_importing.sf}

# This code chunk needs to be written by yourself 
#
# Define sample names 
SampleNames <-  c(paste0("WT-rep", 1:3), paste0("cKO-rep", 1:3)) 

# Define group level
GroupLevel <- c("WT", "cKO")

# Define contrast for DE analysis
Contrast <- c("Group", "cKO", "WT")


# Define a vector for comparing TPM vs Counts effect 
TvC <- c("TPM", "Counts")
levels(TvC) <- TvC

# Define .sf file path
sf <- c(paste0("../salmon_output/", 
               SampleNames,
               ".salmon_quant/quant.sf"))

# Define sample groups
group <- c(rep(GroupLevel[1], 3), rep(GroupLevel[2], 3))

# Create metadata
metadata <- data.frame(Sample=factor(SampleNames, levels=SampleNames),
                       Group=factor(group, levels=GroupLevel),
                       Path=sf)

rownames(metadata) <- SampleNames

# Explore the metadata
print(metadata)
```

## Converting .sf files to txi list 

#### - txi_tpm: stores **TPM** with the argument **"countsFromAbundance="lengthScaledTPM"**

#### - txi_counts: stores **original counts** 

#### - Despite that the TPM matrix is not recommended as an input for DE analysis due to the fact that it doesn't account gene length, TPM stored in a tximport (by **tximport(..., countsFromAbundance="lengthScaledTPM")**) can be used for DE analysis by being inputted with the **DESeqDataSetFromTximport()** funtion in DESeq2 workflow as the gene length is automatically adjusted by DESeqDataSetFromTximport(). 

#### - In this project, two txi objects were created with or without the countsFromAbundance="lengthScaledTPM" argument and compared in downstream DE analysis. 

#### - If you don't want gene-level summarization, set **txOut=TRUE**. 

#### - References: [**tximport doc**](https://bioconductor.riken.jp/packages/3.4/bioc/vignettes/tximport/inst/doc/tximport.html), [**DESeq2 doc "Why unnormalized counts?"**](https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#why-un-normalized-counts), [**Soneson et al. 2016**](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4712774/), [**Developer Dr. Love's comment**](https://support.bioconductor.org/p/84059/), [**Harvard Chan Bioinformatics Core workshop**](https://github.com/hbctraining/DGE_workshop_salmon_online/blob/master/lessons/02_DGE_count_normalization.md)


```{r saving_reads_to_dataframe}

# Assign sample names to the input (.sf) file path
names(sf) <- SampleNames

# Run tximport
# tpm vs original counts
# input sf: a factor of all .sf files' path
txi_tpm <- tximport(sf, 
                    type="salmon",
                    tx2gene=AnnoDb,
                    countsFromAbundance="lengthScaledTPM", # Extracts TPM 
                    ignoreTxVersion=T) 

txi_counts <- tximport(sf, 
                    type="salmon",
                    tx2gene=AnnoDb,
                    ignoreTxVersion=T) 

# tpm 
head(txi_tpm$counts)
dim(txi_tpm$counts)

# counts
head(txi_counts$counts)
dim(txi_counts$counts)
```



## Creating DESeq objects from txi and VST

#### - Note: The tximport-to-DESeq2 approach uses estimated gene counts from the transcript abundance quantifiers, but not normalized counts.

#### - The **DESeqDataSetFromTximport()** function generated an DESeq object (aka **dds**) with the **txi input**. 

#### - **vst()** was run to perform variance stabilizing transformation instead of rlog() which takes longer time with similar characteristics. 

#### - The **vsd** object created by vst() is used for not DE analysis but QC. 

#### - References: [**DESeq2 doc "Transcript abundance files"**](https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#tximport), [**DESeq2 doc "Variance stabilizing transformation"**](https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#variance-stabilizing-transformation)




```{r creating_dds_vsd}

# Set a function creating dds and vsd
dds_vsd_fn <- function(txi) { 

    # Create a DESeq object (so-calledd dds) 
    des <- DESeqDataSetFromTximport(txi, 
                                    colData=metadata,
                                    design=~Group)

    # Create a vsd object (so-called vsd) 
    ves <- vst(des, blind=T)

    # Output them as a list 
    return(list(dds=des, vsd=ves))

}

TPM <- dds_vsd_fn(txi_tpm)
Counts <- dds_vsd_fn(txi_counts)
# Outputs
# dds: TPM/Counts[[1]] or TPM/Counts[['dds']] 
# vsd: TPM/Counts[[2]] or TPM/Counts[['vsd']]
```


## Exploring created dds 

```{r exploring_dds}

# TPM 
TPM[[1]]
head(counts(TPM[[1]]))


# Counts
Counts[[1]]
head(counts(Counts[[1]]))
```

## Exploring created vsd

```{r exploring_vsd}
# TPM
TPM[[2]]
# Counts
Counts[[2]]
```

## Estimating size factors, dispersions, and conducting Wald Test

#### - **Dispersion** is calculated as a **measure of variation** instead of variance since variance gets larger when gene expression gets higher. 

#### - **Wald test** is the default setting of DESeq2 which tests null hypothesis between **two groups**. You should use **Likelihood ratio test (LRT)** when comparing **more than two groups**.     

#### - Messages when "Counts <- DESeqPrep_fn(Counts)" was run: using **'avgTxLength'** from assays(dds)

#### - References: [**Harvard Chan Bioinformatics Core workshop I**](https://github.com/hbctraining/DGE_workshop_salmon_online/blob/master/lessons/05b_wald_test_results.md), [**Harvard Chan Bioinformatics Core workshop II**](https://github.com/hbctraining/DGE_workshop_salmon_online/blob/master/lessons/05a_hypothesis_testing.md), [**Harvard Chan Bioinformatics Core workshop III**](https://github.com/hbctraining/DGE_workshop_salmon_online/blob/master/lessons/04b_DGE_DESeq2_analysis.md), [**DESeq2 "Wald test indivisual steps**](https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#wald-test-individual-steps), [**DESeq2 doc "Likelihood ratio test"**](https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#likelihood-ratio-test)


```{r DESeq_prep}
# Set a function estimating size factors, dispersions, and perform wald test
DESeqPrep_fn <- function(List) {
    
    List[[1]] <- estimateSizeFactors(List[[1]])
    List[[1]] <- estimateDispersions(List[[1]])
    List[[1]] <- nbinomWaldTest(List[[1]])
   
    return(List)
}
# Update dds with the function
Counts <- DESeqPrep_fn(Counts) 
TPM <- DESeqPrep_fn(TPM)
```

## Exploring size factors

```{r exploring_sizefactors}
sizeFactors(Counts[[1]])
sizeFactors(TPM[[1]])
# Size factors don't exist in the Counts dds!
# Normalization factors are calculated in the Counts dds instead! 
assays(Counts[[1]])
assays(TPM[[1]])
colData(Counts[[1]])
colData(TPM[[1]])
```

## Plotting the size factors in TPM

#### - The size factors are only available from TPM dds 

#### - Blue dashed line: normalization factor = 1


```{r plotting_sizefactors}
# Extract and save the size factors in a data frame
sizeFactor <- as.data.frame(round(sizeFactors(TPM[[1]]), 3))

colnames(sizeFactor) <- 'Size_Factor'

sizeFactor <- sizeFactor %>%
    rownames_to_column(var="Sample") %>%
    inner_join(metadata[, 1:ncol(metadata)-1], by="Sample") 

sizeFactor$Sample <- factor(sizeFactor$Sample, levels=SampleNames)



# Create a plot comparing the size factors by sample
ggplot(sizeFactor, aes(x=Sample, 
                       y=Size_Factor, 
                       fill=Group,
                       label=Size_Factor)) +
    geom_bar(stat="identity", width=0.8) +
    theme_bw() + 
    ggtitle("Size Factors in TPM-DESeq") +
    geom_text(vjust=1.5) +
    theme(axis.text.x=element_text(angle=45, 
                                   vjust=0.5)) + ylab("Size Factor") + geom_hline(yintercept=1, color="blue", linetype="dashed")
    
```


## Plotting nornalization factors in the Counts

#### - DESeq2 performs an internal normalization where geometric mean is calculated for each gene across all samples. The counts for a gene in each sample is then divided by this mean. The median of these ratios in a sample is the size factor for that sample.

#### - Blue dashed line: normalization factor = 1

#### - Colored **dots**: **normlization factors per gene** (y-axis) in each sample 

#### - Box plots: distribution of the normalization facters in each group (see the second plot)

#### - Reference: [**DESeq2 doc "Sample-/gene-dependent normalization factors"**](https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#log-fold-change-shrinkage-for-visualization-and-ranking)

```{r plotting_normalizationfactors}
# Extract and normalization factors in a data frame
normf <- as.data.frame(normalizationFactors(Counts[[1]])) %>%
    gather(Sample, Normalization_Factor) %>%
    inner_join(metadata[, 1:2], by="Sample") 
normf$Sample <- factor(normf$Sample, levels=SampleNames)
normf$Group <- factor(normf$Group, levels=GroupLevel)
# Create a scatter plot showing distribution of normalization factors 
normFactors_plot <- ggplot(normf, 
       aes(x=Sample, y=Normalization_Factor)) + 
geom_jitter(alpha=0.5, aes(color=Group)) + 
# Add a boxplot to provide statistics in each sample
geom_boxplot(aes(x=Sample, y=Normalization_Factor), 
             outlier.shape=NA, alpha=0.5) + 
theme_bw() +
ggtitle("Normalization Factors in Counts-DESeq") +
theme(axis.text.x=element_text(angle=45, 
                               vjust=0.5)) + 
ylab("Normalization Factor / Gene") +
# Add a dashed horizontal line to indicate where normalization factor=1
geom_hline(yintercept=1, 
           color="blue", 
           linetype="dashed")
# Print the normalization factor scatter plot 
print(normFactors_plot)
# Print the same plot with larger y-magnification in order to observe the box plot 
normFactors_plot + 
    ylim(0.8, 1.2)
```



## Setting functions for QC plots

#### - Reference: [**DESeq2 doc "Principal component plot of the samples"**](https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#principal-component-plot-of-the-samples), [**DESeq2 doc "Heatmap of the sample-to-sample distances"**](https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#heatmap-of-the-sample-to-sample-distances)

```{r QCplot_functions}

# Assigne what to compare
GroupOfInterest <- Contrast[1] 

# Set a function for a PCA plot
QCPCA_fn <- function(inputList, Title) {

    plotPCA(inputList[[2]],    # takes vsd
            intgroup=GroupOfInterest) + theme_bw() + ggtitle(Title)
}

# Set heatmap annotation 
ColOfInterest <- !colnames(metadata) %in% c("Sample", "Path")
HeatmapAnno <- as.data.frame(metadata[, ColOfInterest])
rownames(HeatmapAnno) <- SampleNames
colnames(HeatmapAnno) <- colnames(metadata)[ColOfInterest]

# Set a function for a correlation heatmap 
QCcorrHeatmap_fn <- function(inputList, Title) {

    # Extract transformed count matrix
    mtx <- assay(inputList[[2]])      # takes vsd

    # Calculate correlation and store in the matrix
    mtx <- cor(mtx)
    
    # Create a correlation heatmap
    return(pheatmap(mtx, 
             annotation=HeatmapAnno,
             main=paste("Sample Correlation Heatmap:",
                        Title)))
}
```


## Sample QC: Principal Component Analysis (PCA)

#### - Checkpoints in PCA: source of variation, sample outlier



```{r QC_PCA}
grid.arrange(QCPCA_fn(TPM, "QC PCA: TPM"), 
             QCPCA_fn(Counts, "QC PCA: Counts"), 
             nrow=2)
```



## Sample QC: Sample Correlation Heatmap

#### - Checkpoints of correlation heatmap: distance between samples, correlation in a group

#### - Upper: TPM input

#### - Lower: Counts input



```{r QC_correlation_heatmap}
# TPM
QCcorrHeatmap_fn(TPM, "TPM") 
# Counts
QCcorrHeatmap_fn(Counts, "Counts") 
```

## Running DE analysis with or without shrinkage

#### - Shrinkage reduces false positives

#### - Magnitude of shrinkage is affected by dispersion and sample size

#### - When the argument **type** is set to **"apeglm"**, the **coef** argument is used instead of constrast. In this dataset, you can set **coef=Coef** where **Coef <- resultsNames(ddsList[[1]])**. 

#### - When the **type** is set to **"normal"**, the argument **contrast** is set as shown below. 

#### - References: [**DESeq2 doc "Alternative shrinkage estimators"**](https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#alternative-shrinkage-estimators), [**Harvard Chan Bioinformatics Core workshop **](https://github.com/hbctraining/DGE_workshop_salmon_online/blob/master/lessons/04b_DGE_DESeq2_analysis.md)

```{r DE_analysis}

# Create a list consisted with dds objects from TPM and Counts
desList <- list(TPM[[1]], Counts[[1]]) 
names(desList) <- TvC

# Create a list for TPM and Counts dds 
ddsList <- desList  # DE without shrinkage  
normal.ddsList <- desList    # DE with "normal" shrinkage
ape.ddsList <- desList       # DE with "apeglm" shrinkage
ash.ddsList <- desList       # DE with "ashr" shrinkage

for (x in TvC) {
    
    # Run DESeq() 
    ddsList[[x]] <- DESeq(desList[[x]])
    print(resultsNames(ddsList[[x]]))

    normal.ddsList[[x]] <- lfcShrink(ddsList[[x]],
                                contrast=Contrast,
                                type="normal")

    ape.ddsList[[x]] <- lfcShrink(ddsList[[x]],
                             coef=resultsNames(ddsList[[x]])[2],
                             type="apeglm")

    ash.ddsList[[x]] <- lfcShrink(ddsList[[x]],
                             coef=resultsNames(ddsList[[x]])[2],
                             type="ashr")

}

# Combine every ddsList into a list
all.ddsList <- list(ddsList, normal.ddsList, ape.ddsList, ash.ddsList)
shrinkage <- c("None", "Normal", "Apeglm", "Ashr")
names(all.ddsList) <- shrinkage
```



## Creating dispersion plots

#### - Dispersion is important since estimation by DESeq2 algorithm is based on the assumption that genes with similar expression levels have similar dispersion. If an RNA-seq dataset doesn't satisfy this assumption, use other DE algorithms than DESeq2. 

#### - References: [**DESeq2 doc "Dispersion plot and fitting alternatives**](https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#dispersion-plot-and-fitting-alternatives), [**Harvard Chan Bioinformatics Core workshop **](https://github.com/hbctraining/DGE_workshop_salmon_online/blob/master/lessons/04b_DGE_DESeq2_analysis.md)



```{r dispersion_plot}

# Plot dispersion  
for (x in TvC) {

    plotDispEsts(ddsList[[x]], 
                 ylab="Dispersion", 
                 xlab="Mean of Normalized Counts", 
                 main=paste("Dispersion of", x, "Input"))
}
```


## Extracting DE results with or without shrinkage

#### - The **alpha** denotes threshold of **false discovery rate (FDR)** assigned by users.

#### - In this analysis, the alpha is set to **0.1** 

```{r extracting_results}

# Set FDR threshold 
alpha=0.1 

# FDR threshold vector
FDRv=c("< 0.1", "> 0.1") 

# Initialize lists of result tables 
all.resList <- all.ddsList 

# Set a function cleaning table
Sig.fn <- function(df, Input) {
    
    df <- df %>% 
        rownames_to_column(var="Gene") %>%
        mutate(FDR=ifelse(padj < 0.1 & !is.na(padj), 
                                   FDRv[1], 
                                   FDRv[2]), 
               Input=Input) 
    return(df)
}




for (i in shrinkage) {

    if (i == "None") {

        for (x in TvC) {

        # Extract data frames from unshrunken lfc & clean data 
        all.resList[[i]][[x]] <- as.data.frame(results(all.ddsList[[i]][[x]], 
                                                       contrast=Contrast, 
                                                       alpha=alpha)) %>% 
        Sig.fn(x)

         } 
    } else {

        # Extract data frames from shrunken lfc & clean data
        for (x in TvC) {

            all.resList[[i]][[x]] <- as.data.frame(all.ddsList[[i]][[x]]) %>%
                Sig.fn(x)
    

        }
    }
}





# Explore the results 
summary(all.resList)
head(all.resList[[1]][['TPM']])
head(all.resList[[1]][['Counts']])
head(all.resList[[2]][['TPM']])
head(all.resList[[2]][['Counts']])
```

## Exploring mean-difference relationship with MA plots


#### - **x-axis**: expression level (baseMean))

#### - **y-axis**: fold change (log2FoldChange)

#### - **Red dashed lines**: log2FoldChange = -1 and 1

#### - Reference: [**DESeq2 doc "MA-plot"**](https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#ma-plot)


```{r MAplot}

# Set ylim: has to adjusted by users depending on data 
yl <- c(-7.5, 7.5)

# Set min log2 fold change of interest 
mLog <- c(-1, 1)

# Initialize a list storing MA plots
MAList <- ddsList


# Create MA plots

for (i in shrinkage) {

    both.df <- rbind(all.resList[[i]][[TvC[1]]], 
                     all.resList[[i]][[TvC[2]]])

    MAList[[i]] <- ggplot(both.df, 
                              aes(x=baseMean, y=log2FoldChange, color=FDR))  +geom_point() + scale_x_log10() + facet_grid(~Input) + 
                                   theme_bw() + 
                                   scale_color_manual(values=c("blue", "grey")) +
                                   ggtitle(paste("MA plot with", i)) + 
                                   ylim(yl[1], yl[2]) + 
                                   geom_hline(yintercept=c(mLog[1], mLog[2]), linetype="dashed", color="red") 

}

   

# Print MA plots
MAList
```


## Exploring distribution of false discovery rate (FDR)

#### - Distribution of adjusted p-val (FDR) was presented 

#### - **x-axis**: FDR

#### - **y-axis**: Number of genes

#### - **Black dashed line**: FDR = alpha

```{r FDR_distribution}


# Subset rows with FDR < alpha
both.df <- rbind(all.resList[[1]][['TPM']], 
                 all.resList[[1]][['Counts']])
both.df$Input <- factor(both.df$Input, levels=TvC)

# Plot distribution of FDR
ggplot(both.df,
       aes(x=padj, color=Input)) + 
geom_density(size=1, aes(y=..count..)) + 
theme_bw() +
ggtitle("Distribution of False Discovery Rate (FDR)") + 
xlab("Adjusted P-Value (FDR)") + 
ylab("Count") + 
geom_vline(xintercept=alpha, 
           color="black", 
           linetype="dashed",
           size=1) + 
scale_x_continuous(breaks=seq(0, 1, by=0.1)) 


```


## Exploring distribution of log2FoldChange by input type

#### - **Black dashed lines**: log2FoldChange = -1 and 1 

#### - **x-axis**: gene expression level (log2FoldChange) 

#### - **y-axis**: number of genes 

```{r L2FC_distribution}
# Initialize a lfc list
lfcplotList <- all.resList 

# Create lfc distribution plots
for (i in shrinkage) {

    lfc.df <- rbind(all.resList[[i]][[TvC[1]]], 
                    all.resList[[i]][[TvC[2]]])

    lfc.df <- lfc.df[lfc.df$FDR == "< 0.1",]

    lfc.df$Input <- factor(lfc.df$Input, levels=TvC)

    lfcplotList[[i]] <- ggplot(lfc.df,  # Subset rows with FDR < alpha
                               aes(x=log2FoldChange, color=Input)) + 
geom_density(size=1, aes(y=..count..)) + 
theme_bw() + ylab("Count") + 
geom_vline(xintercept=c(mLog[1], mLog[2]), 
           color="black", 
           linetype="dashed", 
           size=1) + 
ggtitle(paste("Distribution of Log2FoldChange by Input Type:", i)) + 
xlim(-5, 5)


}

# Print the lfc plots
lfcplotList
```



## NA statistics: zero count genes & outlier genes  


#### When NAs appear in  

#### - **log2FoldChange**: zero counts in all samples

#### - **padj**: too little information 

#### - **pval & padj**: at least one replicate was an outlier 



```{r DEG_statistics}

# Count number of NA genes  
type=c("Zero Counts", "Outliers", "Total NA Genes") 


# Create a data frame storing NA gene number
NAstat <- both.df %>%
    group_by(Input) %>%
    summarize(zero=sum(is.na(log2FoldChange)), 
              outlier=sum(is.na(pvalue) & is.na(padj))) %>%
    mutate(total=zero + outlier) %>%
    gather(Type, Number, -Input) %>%
    mutate(Type=factor(case_when(Type == "zero" ~ type[1], 
                                 Type == "outlier" ~ type[2], 
                                 Type == "total" ~ type[3]), 
                       levels=type))

# Plot number of NA genes 
ggplot(NAstat, 
       aes(x=Type, y=Number, group=Input, fill=Input, label=Number)) + 
    geom_bar(stat="identity", position="dodge") + 
    theme_bw() +
    geom_text(position=position_dodge(width=1), vjust=1.5) + 
    ggtitle("Number of NA Genes") + 
    ylab("Number of Genes")
```


## baseMean/LFC/FDR comparison between TPM and Count inputs



```{r calculate_simple_difference}


# Create a data frame storing the number of transcripts by gene id
AnnoDb.ntrans <- AnnoDb %>% 
    group_by(GENEID) %>% 
    summarize(num.trans=n_distinct(TXID))

# Create an empty list storing significant gene lfc tables 
sigList <- list() 


# Filter significant genes' lfc and save in the list 
for (x in shrinkage) {


    for (y in TvC) {

        # Subset genes with FDR below alpha
        sigList[[x]][[y]] <- subset(all.resList[[x]][[y]], 
                                    FDR == paste("<", alpha))

        # Explore the output 
        print(head(sigList[[x]][[y]]))
        print(dim(sigList[[x]][[y]]))

    }
}


# Clean the data frames with renaming columns
for (x in shrinkage) {

    # Join TPM and Counts tables by GENEID
    df <- inner_join(sigList[[x]][[1]], 
                     sigList[[x]][[2]],
                     by="Gene") 

    # Create a vector storing original column nams 
    my.colname1 <- colnames(sigList[[x]][[1]])[-1]

    # Create a vector storing modified column names
    my.colname2 <- c("GENEID", 
                     paste0(my.colname1, "_", TvC[1]), 
                     paste0(my.colname1, "_", TvC[2]))

    # Rename the columns
    colnames(df) <- my.colname2

    # Add a column storing shrinkage method and drop redundant columns
    df <- df %>% 
        mutate(Shrinkage = x) %>% 
        dplyr::select(-starts_with(c("lfcSE", "stat", "FDR")))

    # Save the cleaned data frame in the list
    sigList[[x]] <- df

    # Explore the output data frame
    print(head(df))
    print(dim(df))
}

# Convert a list of data frames to one single data frame 
lfcTable <- sigList[[1]] 

for (i in 2:length(shrinkage)) {

    lfcTable <- rbind(lfcTable, sigList[[i]])

}

# Explore the output data frame
head(lfcTable)
dim(lfcTable)


# Calculate differences between TPM and Counts input data 
# in baseMean, log2FoldChange, and padj
lfcTable <- lfcTable %>%

    mutate(mean_TC=baseMean_TPM - baseMean_Counts,
           lfc_TC=log2FoldChange_TPM - log2FoldChange_Counts,
           FDR_TC=padj_TPM - padj_Counts) %>%

# Add a column storing the number of alternative transcripts
left_join(AnnoDb.ntrans, by="GENEID")

# Explore the output data frame
head(lfcTable)
dim(lfcTable)

# Set a function to create a vector storing plot labels
plotlabels.fn <- function(myvec, mylist, num) {

    vec <- c()

    for (i in 1:num) {
        vec <- c(vec, c(myvec[i], rep("", nrow(mylist[[i]]) - 1)))
    }

    return(vec)

}




```


```{r calculate_correlation}

my.param <- c("baseMean", "log2FoldChange", "padj")


# Slice and clean the data frame for input
lfcTable.comp <- lfcTable %>% 
    dplyr::select(GENEID, num.trans, Shrinkage, starts_with(my.param)) %>%  
    gather(Category, Value, -GENEID, -num.trans, -Shrinkage) %>% 
    separate(Category, c("Metric", "Input"), sep="_") %>% 
    pivot_wider(names_from=Input, values_from=Value) %>%
    nest(-Metric, -Shrinkage)  

# Create a vector storing Rsquared values between TPM and Counts outputs
corr.vec <- round(map_dbl(lfcTable.comp$data, 
                          ~cor(.x$TPM, .x$Counts)),
                  7)

# Create a ggplot labeling vector converted from the Rsquared vector
rsq.vec <- plotlabels.fn(corr.vec, 
                         lfcTable.comp$data, 
                         length(corr.vec))


# Unnest the data frame
lfcTable.comp <- lfcTable.comp %>% 
    unnest(data) 

# Add a column storing Rsquared labels 
lfcTable.comp$Rsquared.label <- rsq.vec


# Explore the cleaned data frame
head(lfcTable.comp)

# Nest the data frame by metric
lfcTable.comp <- lfcTable.comp %>%
    nest(-Metric) 



```


```{r display_comparison_plots, fig.width=10, fig.height=10} 

# Set a function creating a scatter plot
comp.scatter.fn <- function(df, 
                            xvar, 
                            yvar, 
                            met, 
                            xlabel, 
                            ylabel) {

    p <- ggplot(df, aes(x=xvar, 
                        y=yvar, 
                        color=log(num.trans), 
                        label=Rsquared.label)) + 
        geom_point(alpha=0.5) + 
        theme_bw() + 
        ggtitle(paste("Comparison in", met, "\n(with R-Squared)")) + 
        geom_text(size=5, 
                  mapping=aes(x=Inf, y=Inf), 
                  vjust=2, hjust=3.8, color="black") +
geom_abline(slope=1, size=0.5, linetype="dashed", color="black") + 
scale_color_gradient(low="blue", high="red") +
facet_wrap(~Shrinkage, scales="free") + 
theme(strip.text.x=element_text(size=10)) +
xlab(paste("Input from", xlabel)) + 
ylab(paste("Input from", ylabel))


return(p)
}




# Print the plots
for (i in 1:nrow(lfcTable.comp)) {

    df <- lfcTable.comp$data[[i]]

    print(comp.scatter.fn(df,
                          df$TPM,
                          df$Counts,
                          lfcTable.comp$Metric[i], "TPM", "Counts"))
    
}



```


## baseMean/LFC/FDR  rank comparison between TPM and Count inputs



```{r rank_comparison, fig.width=10, fig.height=10}

# Transform the data frame
lfcTable.rank <- lfcTable.comp %>% 
    unnest(data) %>%
    nest(-Metric, -Shrinkage)

# Clean and arrange the data frame
for (i in 1:length(lfcTable.rank$data)) {

    df <- lfcTable.rank$data[[i]] 

    # Create a vector storing rank (1 to the last)
    rank.vec <- 1:nrow(df)

    # Make descending order if the metric = "padj" and assigne rankings
    if (lfcTable.rank$Metric[[i]] == "padj") {

        df <- df %>%
            dplyr::arrange(TPM) %>% 
            mutate(rank.TPM=rank.vec) %>%
            dplyr::arrange(Counts) %>%
            mutate(rank.Counts=rank.vec) 

    # Make asending order otherwise and assign rankings
    } else {

        df <- df %>%
            dplyr::arrange(desc(abs(TPM))) %>%
            mutate(rank.TPM=rank.vec) %>% 
            dplyr::arrange(desc(abs(Counts))) %>%
            mutate(rank.Counts=rank.vec)

    }



    # Save the updated data frame 
    lfcTable.rank$data[[i]] <- df 



}

# Create a vector storing Rsquared values between TPM and Counts outputs
corr.vec <- round(map_dbl(lfcTable.rank$data, 
                          ~cor(.x$rank.TPM, .x$rank.Counts)),
                  7)

# Create a ggplot labeling vector converted from the Rsquared vector
rsq.vec <- plotlabels.fn(corr.vec, lfcTable.rank$data, length(corr.vec))

# Unnest the data frame
lfcTable.rank <- lfcTable.rank %>% 
    unnest(data) 

# Add a column storing Rsquared labels 
lfcTable.rank$Rsquared.label <- rsq.vec


# Explore the cleaned data frame
head(lfcTable.rank)

# Nest the data frame by metric
lfcTable.rank <- lfcTable.rank %>%
    nest(-Metric) 




# Print the plots
for (i in 1:nrow(lfcTable.rank)) {

    df <- lfcTable.rank$data[[i]]


    pl <- comp.scatter.fn(df,
                          df$rank.TPM,
                          df$rank.Counts,
                          lfcTable.rank$Metric[i], 
                          "TPM", "Counts") + 
ggtitle(paste("Comparison in", lfcTable.rank$Metric[i], "Rank\n(with R-Squared)")) 

print(pl)

 
}

```



## Summarizing up/down DEGs with an upset plot 

#### - **red bar**: input type

#### - **blue bar**: directionality of gene expression change

```{r upsetplotting, fig.width=10}

# Set a function cleaning data to generate upset plots 
upset.input.fn <- function(df) {

    df <- df %>% 

        # Filter genes with valid padj 
        subset(!is.na(padj)) %>% 
        
        mutate(Up=ifelse(FDR == FDRv[1] & log2FoldChange > 0, Gene, ""), # What are upregulated genes? 
               
               Down=ifelse(FDR == FDRv[1] & log2FoldChange < 0, Gene, ""),  # What are downregulated genes? 
               
               Unchanged=ifelse(FDR == FDRv[2], Gene, ""),   # What are unchanged genes? 
               
               TPM_Input=ifelse(Input == "TPM", Gene, ""),   # What are the genes from TPM input? 
               
               Counts_Input=ifelse(Input == "Counts", Gene, ""))   # What are the genes from Counts input? 

    # Create a list storing groups of interest
    upsetInput <- list(Up=df$Up, 
                       Down=df$Down, 
                       Unchanged=df$Unchanged, 
                       TPM_Input=df$TPM, 
                       Counts_Input=df$Counts) 

    return(upsetInput)

}

upsetList <- upset.input.fn(both.df)


# Create the upset plot 
upset(fromList(upsetList), 
      sets=names(upsetList),   # What group to display 
      sets.x.label="Number of Genes per Group",
      order.by="freq",
      point.size=3,
      sets.bar.color=c("red", "red","blue", "blue", "blue"),
      text.scale = 1.5, number.angles=30) 
```



## Session Info 

```{r sessionInfo}
sessionInfo()
```
